---
title: "Water Quality and Spatial Data Preparation"
author: "Matthew Ross"
date: "3/15/2017"
output: html_document
---


In this lesson, we will look at how land use change affects downstream water quality. 

#Data Prep
```{r,echo=F,include=F}
#Read in key libraries 
library(rgdal) #Geospatial library for open source spatial analyses
library(rgeos) #Database of projeciton transformations
library(raster) #Library for importing and analyzing rasters
library(tidyverse) #Metalibrary that pulls in key data manipulation libraries
library(reshape2) #Data manipulation library
library(leaflet) #Data visualization of spatial data
library(lubridate) #Easy way to assign time stamps
library(dataRetrieval) #Package that direclty taps into the USGS database
```

##Read in raw data 

Data comes from several sources

* Q data comes from USGS for stations: 
    + 09046600, Blue River above Dillon Reservoir
    + 09047500, Snake River near Montezuma
* Water Quality is downloaded from STORET  [database](https://ofmpub.epa.gov/apex/STORETSummary/f?p=WATERSHEDUI:1:::::P1_ORG_CHAR,P1_HUC:1,14010002) from several sources:
    + RiverWatch Network
    + Colorado Department of Publich Health and Environment
* Spatial data comes from [NHD Plus](http://www.horizon-systems.com/nhdplus/)


```{r,echo=F,eval=F}
#Read in watershed boundary dataset and subset to blue river watershed station data
catch <- readOGR('data/NHDPlusCO-1/NHDPlus14/WBDSnapshot/WBD','WBD_Subwatershed') %>%
  .[.$HUC_8 == 14010002,]


#Read in and subset STORET data  from the 3 different sources
riverwatch <- read.delim('data/riverwatch/Data_rvm_20170315_122429_RegResults.txt',sep='\t',stringsAsFactors = F)  

cdphe.new <- read.delim('data/cdphe.new/Data_vrm_20170315_122407_RegResults.txt',sep='\t',stringsAsFactors = F) 
#  filter(Station.ID %in% c('12336','12337'))

cdphe.old <- read.delim('data/cdphe.old/chem.txt',sep='\t',stringsAsFactors = F) 
#  filter(Station.ID %in% c('12336','12337'))

#Bind all 3 of those watershed data sources
dat.full <- rbind(cdphe.new,cdphe.old,riverwatch)

#Identify Snake River and Blue River monitoring stations
keepers <- c('000115','000140','51')
stat.names <- c('Blue River', 'Snake River','Snake River')

#Cast and filter data by number of observations to > 100
dat.cast <- dat.full %>%
  group_by(Station.ID, Station.Longitude,Station.Latitude) %>%
  summarise(unq = length((unique(Activity.Start)))) %>%
  filter(unq>100) %>% 
  arrange(desc(unq)) %>%
  filter(Station.ID %in% keepers)


#Convert cast data into a spatial data frame
sp.dat <- dat.cast %>% as.data.frame(.)
station.sp <-  SpatialPointsDataFrame(coords = sp.dat[,c('Station.Longitude','Station.Latitude')], data = sp.dat, proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")) %>%
  spTransform(.,CRS=projection(catch))


#Subset spatial data to only the huc10 watersheds of interest
csub <- catch[catch$HU_10_NAME %in% c('Snake River','Upper Blue River'),]



#Subset full chemistry dataset to only Upper Blue River and Snake River
chem.dat <- dat.full %>% 
  filter(Station.ID%in%keepers) %>%
  mutate(Site=ifelse(Station.ID==keepers[1],'Upper Blue River','Snake River'))

#Read in Q data. 
q.dat <- readNWISdata(service='dv',sites=c('09046600','09047500'),
                        parameterCd='00060',
                        startDate='1970-05-01',endDate='2016-10-10')

#Add in sensible site names
q.dat$Site = ifelse(q.dat$site_no=='09046600','Upper Blue River','Snake River')


#Save full unmerged datasets
save(q.dat,chem.dat,csub,file='data/UnmergedDat.RData')
```



##Manipulate and reshape data to extract yearly means of concentration data for key constituents. 
```{r, echo=F}

load('UnmergedDat.RData')

#List of analytes of interest
key.params <- c('Aluminum','Cadmium','Copper','Calcium','Inorganic nitrogen (nitrate and nitrite) as N','Phosphate-phosphorus as P',
                'Iron','Lead','Manganese','Magnesium','pH','Selenium','Sodium','Zinc','Specific conductance','Sulfur, sulfate (SO4) as SO4')


#Subset data by a few key characteristics and convert all data to mg/l 
#Note that pH has no units and SC is in uS/cm
sub.full <- chem.dat %>%
  filter(Characteristic.Name %in% key.params) %>%
  mutate(Units=ifelse(grepl('mg/l',Units),'mg/l',Units)) %>%
  mutate(Units=ifelse(grepl('ug/l',Units),'ug/l',Units)) %>%
  mutate(Result.Value.as.Number=ifelse(Units=='ug/l',Result.Value.as.Number/1000,Result.Value.as.Number)) %>%
  mutate(Units=gsub('ug/l','mg/l',Units)) %>%
  mutate(Units=ifelse(grepl('None',Units),'None',Units)) %>% 
  mutate(Analyte = Characteristic.Name) %>%
  mutate(date=ymd_hms(Activity.Start)) %>%
  filter(date> ymd_hms('1970-01-01 00:00:00')) %>%
  mutate(dateTime=as.Date(date))


#Reshape data to make it easier for analysis
chem.cast <- dcast(sub.full,dateTime+Site ~ Analyte, value.var='Result.Value.as.Number',mean)
#Melt data back into long format
chem.melt <- melt(chem.cast,id.vars=c('dateTime','Site'),measure.vars=key.params) %>% filter(!is.na(value))

#Join chem data to full q datasets
#First have to convert posix to date object
q.dat$dateTime <- as.Date(q.dat$dateTime)
q.chem <- left_join(q.dat,chem.melt, by=c('Site','dateTime'))


#Save this q.chem dataset with spatial data for shiny app
csub$color=c('blue')
csub$color[csub$HU_10_NAME == 'Snake River'] <- 'red'

#Name discharge column by untis (cubic feet per second)
names(q.chem)[4] <- 'cfs'
#Convert to lps and m3s
q.chem$Lps <- q.chem$cfs*28.3168
q.chem$m3s <- q.chem$Lps/1000
q.chem$variable <- as.character(q.chem$variable)
#Cast by site
q.cast <- dcast(q.chem,dateTime~Site,value.var='m3s',mean)
#Convert to time series object
q.xts <- xts(q.cast[,-1],order.by=q.cast$dateTime)


save(q.chem,q.cast,csub,station.sp,file='data/ShinyFullDataset.RData')

```

#Data Visualization
SHINY APP

#Landuse change
12sV8yVPaWm9BipuXrikKNweB297mi4SQAPrCUxms

[Climate](http://clim-engine.appspot.com/climateEngineExpert?toolAction=getMap&productType=RS&product=L_SR&variable=NDVI&statistic=Mean&calculation=anompercentchange&units=metric&varUnits=%&dateStart=2016-06-16&dateEnd=2016-09-16&yearStart=1982&yearEnd=1990&yearStartClim=1984&yearEndClim=1990&opacity=0.7&mapCenterLongLat=-105.8968,39.5483&mapzoom=13&timeperiod=custom&minColorbar=-30&maxColorbar=30&colorbarmap=RdYlGn&colorbarsize=8&colorbarType=continuous&layer=custommaplayer&palette=d73027,f46d43,fdae61,fee08b,d9ef8b,a6d96a,66bd63,1a9850&basemap=hybrid)
```{r}

```


